---
title: "The Problem With Historical Instrumental Variables"
author: "John T.H. Wong"
format: 
  pdf:
    documentclass: article
    papersize: letter
    fontsize: 12pt
    geometry:
      - margin=1in
    link-citations: true
    include-in-header: 
      text: |
        \usepackage{tikz}
        \usetikzlibrary{arrows,shapes,positioning,shadows,trees}
    fig-pos: 'H'
    tbl-pos: 'H'
bibliography: references.bib
execute:
  echo: false
  message: false
  warning: false
number-sections: true
---

```{r}
figsize = "80%"
```

I argue in this essay that many papers that use historical (i.e., time-invariant) are misspecified. This is because if the instrument influences the treatment, it must also influence the lags of treatment. This paper makes two contributions. First, using Monte Carlo methods, I prove that two-stage least squares (2SLS) estimates are not only biased and but also inconsistent. Second, I show that a consistent 2SLS estimator of the *instantaneous* treatment effect can be obtained by simply including lagged treatment *and*/or lagged outcome as control variables (a.k.a. exogenous regressors). The simplicity stands in contrast to [bib]'s nonlinear two-parameter estimator, which requires discretion in choosing the initial shock year and requires the delta method for standard error estimates, only to recover a long-run treatment effect. Unlike proxy SVARs (see [bib] for a detailed explainer), the proposed setup also does not require a time-variant instrument---which is not available when attempting to identify a treatment over long periods of time. However, the proposed setup does require a somewhat long panel to obtain unbiased estimates.

This essay is structured as follows. @sec-adl demonstrates, in the simple case of a treatment that enters the outcome equation as an AR(1) sequence, how the 2SLS estimator is inconsistent and the proposed correction. @sec-var extends the analysis to a case of bivariate Granger causality. @sec-consistency uses Monte Carlo to infer the necessary panel length and width under which the correction is robust. @sec-conclude concludes.

## The AR(1) Treatment Case

@fig-dag-adl illustrates how most instrumental variable papers violate exclusion restriction. The exclusion restriction requires that $z_i$ affect $y_{it}$ only through $d_{it}$. Take [bib] for example, who use settlers mortality ($z_i$) in former colonies to identify the effect of expropriation risk in $(d_{it},t = [1985,1995])$ on log output per capita in $(y_{it}, t = 1995)$. For a thorough list of papers that use a similar design, see [bib]. 

::: {#fig-dag-adl fig-env="figure*"}

\begin{tikzpicture}[node distance=2cm]

  % Nodes
  \node (z)    {$z_{i}$};
  \node (dtm1)  [right=of z] {$d_{i,t-1}$};
  \node (dt)    [right=of dtm1] {$d_{it}$};
  \node (yt)    [below=of dt] {$y_{it}$};

  % Arrows
  \draw[->] (z) -- (dtm1);
  \draw[->, bend left=45] (z) to (dt);
  \draw[->] (dtm1) -- (dt);
  \draw[->] (dtm1) -- (yt);
  \draw[->] (dt) -- (yt);

\end{tikzpicture}

Directed acyclic graph: lagged treatment determines treatment and outcome.

:::

The problem with this identification strategy is that if $z_i$ is correlated with $d_{it}$, then it should also be correlated $d_{i, t -1}$, or $d_{i, t-j}$ for up till the $j$ lags of $d_{it}$ that enter into the function for $y_{it}$.

Let us stipulate that only the first lag of the treatment determines the outcome. More formally, suppose $y_{it}$ and $d_{it}$ are respectively determined by the following equations:

$$
\begin{aligned}
y_{it} = \beta_0 d_{it} + \beta_1 d_{i, t-1} + \epsilon_{y,it}
\\ d_{it} = \delta z_i + \alpha_0 d_{i,t-1} + \epsilon_{d, it},
\end{aligned}
$$

where $\epsilon_{y, it}$ and $\epsilon_{d, it}$ are iid. I set intercepts in both equations to zero without loss of generality. I assume both equations are stationary.

The consequence of omitting $d_{i, t-1}$ can be analyzed by solving the $d_{it}$ by iteration, which yields:

$$
d_{it} = (\delta  \sum_{j =0}^{\infty} a_1^j)z_{i}+\sum_{j=0}^{\infty} a_1^j \epsilon_{i,t-j}.
$$

There are two potential misspecifications here. First, the parameter on $z_i$ will be biased by $\delta  \sum_{j =0}^{\infty} a_1^j - \delta = \delta \sum_{j =1}^{\infty} a_1^j$, though this is perhaps not an issue as the first-stage results are not necessarily of interest to researchers. The second issue is that error term potentially violates white noise assumptions required for consistent estimates. If each unit $i$ is only observed for one period, then $d_{it} \rightarrow d_i$, and the term is still independent across units. However, if a panel of $d_{it}$ is used, then the error of more recent observations will be a sum of past errors, violating the assumption of white noise.

Perhaps more concerining is that our estimator of $\beta_0$ in the second-stage equation will be biased. For this I can analyze the 2SLS estimator:

$$
\beta_{\text{2SLS}} = \frac{\mathrm{Cov}(y_{it}, z_{i})}{\mathrm{Cov}(d_{it}, z_{i})}
\\ = \beta_0 + \beta_1 \frac{\mathrm{Cov}(d_{i,t-1}, z_{i})}{\mathrm{Cov}(d_{it}, z_{i})}.
$$

Notice that if there are a long number of periods between $t$ and whenever $z_i$ is determined, $\mathrm{Cov}(d_{it}, z_i) = \mathrm{Cov}(d_{i,t-1}, z_i) = \delta/(1-\alpha_1)$. In this case, the estimator yields the biased result:

$$
\beta_{\text{2SLS}} = \beta_0 + \beta_1.
$$

I can further demonstrate this point by simulating a panel of observations that are determined by the true relationships. The simulated panel has 50 units, each with 1000 periods. Column 1 of @tbl-adl shows the true parameters. When estimated with a naive 2SLS setup, the parameters are biased (Column 2). 


```{r tbl-adl}
#| results: asis
#| tbl-cap: "Two-stage least squares, second-stage results"

source("code.R")
table_lagd

```

I then use Monte Carlo methods to repeat the simulation for 500 iterations. Each iteration still has 50 units, but now with only 100 periods each. @fig-biased shows that the estimated coefficients are consistently biased from the true value, in the direction our analysis predicts.

```{r fig-biased}
#| fig-cap: "Monte Carlo Results, omitted treatment lag (500 iterations; 50 units; 100 observations per unit; ± 2 SD shaded; black dotted line indicates true mean)"

plot_biased
```


I propose a simple correction method: we include $d_{t-1}$ as a control variable. The results are indicated in @tbl-adl, Column 3. Note that this setup is able to recover the true parameters, even though I did not need to use an additional instrument to identify the lagged treatment. Using Monte Carlo, I can show that this specification robustly recovers the true parameter value of $\beta_0$ across random samples (@fig-adl).

```{r fig-adl}
#| fig-cap: "Monte Carlo Results, with treatment lag (500 iterations; 50 units; 100 observations per unit; ± 2 SD shaded; black dotted line indicates true mean)"

plot_labrecque
```

## Adding Bi-Directional Granger Causality

Most dynamic systems present an additional challenge: the contemporaneous treatment is determined by the lags of the outcome. @fig-dag-var builds on @fig-dag-adl, with the additional relationships illustrated with dotted lines. Notice that $y_{i, t-1}$ feeds into both $d_{it}$ and $y_{it}$. Note that I posit only unidirectional contemporaneous causality: $d_{it}$ affects $y_{it}$ (and $d_{i,t-1}$ affects $y_{i,t-1}$), but not the other way around. This type of treatment is found in the growth literature for example; capital in the Solow model depends on past capital and output, but not contemperaneous output.

::: {#fig-dag-var fig-env="figure*"}

\begin{tikzpicture}[node distance=2cm]

  % Nodes
  \node (z)    {$z_{i}$};
  \node (dtm1)  [right=of z] {$d_{i,t-1}$};
  \node (dt)    [right=of dtm1] {$d_{it}$};
  \node (yt)    [below=of dt] {$y_{it}$};
  \node (ytm1)  [below=of dtm1] {$y_{i,t-1}$};

  % Arrows
  \draw[->] (z) -- (dtm1);
  \draw[->, bend left=45] (z) to (dt);
  \draw[->] (dtm1) -- (dt);
  \draw[->] (dtm1) -- (yt);
  \draw[->] (dt) -- (yt);
  \draw[->, dashed] (dtm1) -- (ytm1);
  \draw[->, dashed] (ytm1) -- (dt);
  \draw[->, dashed] (ytm1) -- (yt);
\end{tikzpicture}

Directed acyclic graph: lagged treatment determines lagged outcome and present treatment and outcome. Lagged outcome determines present treatment and outcome.

:::

We can more compact represent the implied sets of equations with a VAR system (even though we do not estimate it as one):

$$
\begin{aligned}
\begin{bmatrix} 1 & -\beta \\ \color{red}{0} & 1\end{bmatrix} \begin{bmatrix} y_{it} \\ d_{it} \end{bmatrix} = \begin{bmatrix} \alpha_{11} & \alpha_{12} \\ \alpha_{21} & \alpha_{22} \end{bmatrix} \begin{bmatrix} y_{i, t-1} \\ d_{i,t-1}\end{bmatrix} + \begin{bmatrix} \color{red}{0} \\ \delta \end{bmatrix} z_i + \begin{bmatrix}\epsilon_{y,it} \\ \epsilon_{d,it}\end{bmatrix}.
\end{aligned}
$$

Several features are immediately apparent:

1. The unidirectional causality from $d_{it}$ to $y_{it}$ in @fig-dag-var is analogous to a Cholesky decomposition that stipulates the treatment as the more exogenous variable.
2. The omission of $z_{i}$ in the second-stage is analogous to forcing the coefficient of the instrument in the first row to be zero in a VAR system.
3. The $\alpha_{ij}$ parameters allow for the estimation of Granger causality. Historical IV papers that omit the $A$ matrix are essentially imposing $\alpha_{ij} = 0 \ \forall  \ i , j$, which is arguably an onerous set of restrictions.

Again, I simulate a panel of observations, and then attempt to recover the true parameters. @tbl-var displays the results. The left-most sub-columns show the true parameters. Columns (a) and (c) show results from a naive 2SLS procedure, whereas Columns (b) and (d) are results from a 2SLS procedure where lagged treatment and outcome are included in the second-stage *and the first-stage* equations. 

```{r tbl-var}
#| results: asis
#| tbl-cap: "Two-stage least squares results, comparison"

table_solow

```

As before, the naive estimates are significantly biased, whereas the proposed specification can recover the true estimates. These results are robust in Monte Carlo (@fig-var).

```{r fig-var}
#| fig-cap: "Monte Carlo results, with treatment and outcome lag (500 iterations; 50 units; 100 observations per unit; ± 2 SD shaded)"

plot_unbiased
```

\newpage

## References